\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-08-01 NAR_DRAFT.tex
add substitution type

At every site in the genome with non-zero coverage, Next Generation Sequencing (NGS) produces a vector $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots D$ of base calls and their associated quality scores, where $D$ is total read depth.
We want to use $\mathbf{x}$ to select between competing hypotheses;
$$
  \begin{array}{l}
    \mathbf{H_0}:\quad \textrm{Alt allele} = m;\quad\nu = 0\\
    \mathbf{H_1}:\quad \textrm{Alt allele} = m;\quad\nu = \hat{f},
  \end{array}
$$
where $\nu$ is the variant allele frequency, $\hat{f}$ is the maximum likelihood estimate of $\nu$ given data $\mathbf{x}$, i.e. the ratio of the count of variant reads and total read depth, and $m$ is any of the 3 possible alternate non-reference bases.
The posterior probability of a given hypothesis, $P(m,\nu)$, is the product of the likelihood of the data given the hypothesis and the prior probability of the hypothesis. 
Assuming that reads are independent, we have

\begin{equation}
  \label{eqn:1}
$$
  P(m,\nu) = p(m,\nu) \cdot \prod_{i=1}^{D} \textrm{f}_{m,\nu}(x_i),
$$
\end{equation}
where the density $\textrm{f}_{m,\nu}(x_i)$ can take several forms.

@@

\subsection{Estimation of the mutation profile}
There are many aspects of genomic architecture which can affect the somatic mutation rate at multiple scales \cite{Buisson2019}.
Here we focus on a small-scale feature, the tri-nucleotide context, which has been shown to have a strong effect on the prior probability of mutation \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
The tri-nucleotide context at a particular base is made up of the identity of the reference base, the identity of the alternate allele, and the identity of the 3` and 5` flanking bases.
There are four possible bases {\{A,C,T,G\}} 3` and 5` of the site, leading to 16 possible flanking base combinations.
Folding the central bases to the pyrimidines leaves 2 possible reference bases at the site {\{C,T\}}, and three possible alternate alleles {\{A,(C/T),G\}} for each.
Thus, there are $2 \cdot 3 \cdot 16 = 96$ possible combinations of flanking bases, reference bases, and alternate alleles.

\begin{figure}[t]
  \begin{center}
  \includegraphics{figures/signature_figure.pdf}
  \end{center}
  \caption{Mutation profiles. 
  \textbf{(a)} The observed mutation profile of the AML31 primary tumor. 
  \textbf{(b)} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
  \textbf{(c)} Equal proportions of signatures 1, 4, \& 5.
  \textbf{(d)} Equal proportions of signatures 1, 3, \& 5.
  KL divergence between the estimated profile ($\boldsymbol{\pi}$) and the simulated mutation profile for \textbf{e)} Whole exomes and \textbf{f)} whole genomes.
  The estimated mutation profile converges quickly to the target as the number of mutations evaluated rises.
  }
  \label{NAR-sigfig}
  \end{figure}

Figure \ref{NAR-sigfig} \textbf{a-d} shows several examples of mutation profiles from both real and simulated tumors.
The mutation profiles in Figure \ref{NAR-sigfig} can be described compactly by the vector $\boldsymbol{\pi} = \{\pi_1,\dots,\pi_{96}\}$ where each element of $\boldsymbol{\pi}$ represents the proportion of all mutations present in the tumor that occur at each of the 96 possible contexts.
We define a mutation context $C$, indexed by $i=\{1 \dots 16\}$ flanking base combinations, $j = \{1,2\}$ reference bases, and $k = \{1,2,3\}$ alternate alleles. 
The distribution of observed contexts $ p(C \mid M) $, where $M$ is the condition that a mutation has occurred, is multinomial with parameter $ \boldsymbol{ \pi } = \{\pi_{i,j,k}\} $.
In a tumor with a very large number of verified mutations $\boldsymbol{\pi}$ could be estimated directly from the observed mutation profile, but in practice we find that many elements of $\boldsymbol{\pi}$ will have $0$ or a very small number of mutations present in the tumor.
As a result we need to model $p(C \mid M)$ as dirichlet - multinomial with pseudo-count hyper-parameter $\boldsymbol{\alpha}$,

$$
\begin{aligned}
  \boldsymbol{\pi} \mid \boldsymbol{\alpha} &\sim \textrm{Dirichlet}(\boldsymbol{\alpha}) \\
  C \mid M,\boldsymbol{\pi} & \sim \textrm{Multinomial}(\boldsymbol{\pi}).
\end{aligned}
$$
In \batcave we use the symmetric non-informative hyper-parameter $\boldsymbol{\alpha} = \boldsymbol{1}$, so a-priori a mutation is equally likely in any context.

@@

Let $\mathrm{D}$ be the set of high confidence variant calls, and $c \in \mathrm{D}$ be the context of each mutation in $\mathrm{D}$.
The posterior distribution of $\boldsymbol{\pi}$ is then $p(\boldsymbol{\pi} \mid D) \sim \textrm{Dirichlet}(\boldsymbol{\alpha^{\prime}})$ where
  $$
    \alpha^{\prime}_{i,j,k} = \alpha_{i,j,k} + \sum\limits_{c \in \mathrm{D}} \mathrm{I}\{c = c_{i,j,k}\}.
  $$
Returning to equation \ref{eqn:full_prior}, we can compute the posterior predictive probability $p(C_{i,j} \mid M,\mathrm{D})$ as
\begin{equation}
  \label{eqn:post_pred}
  $$
  p(C = C_{i,j} \mid M,\mathrm{D}) = \frac{\sum_{k}\alpha^{\prime}_{i,j,k}}{\sum_{i,j,k}\alpha^{\prime}_{i,j,k}}.
  $$
\end{equation}
The posterior probability of mutation to allele $m$ given that a mutation has occurred in context $C_{i,j}$, $p(m \mid C_{i,j},\mathrm{D})$ is then
\begin{equation}
  \label{eqn:to_allele}
  $$
   p(m = k \mid C_{i,j},\mathrm{D}) = \frac{\alpha^{\prime}_{i,j,k}}{\sum_{k} \alpha^{\prime}_{i,j,k}}.
  $$
\end{equation}
The remaining context dependent component in equation \ref{eqn:full_prior} is $p(C)$, the prior probability of seeing a mutation with a particular tri-nucleotide context.
We compute this as simply the proportion of all tri-nucleotide contexts with the context $C_{i,j}$ divided by three. \bkmcomment{I have a pre-computed table in batcaver for wes and wgs. Should I add that bit here?}

>>>>

At every site in the genome with non-zero coverage, Next Generation Sequencing (NGS) produces a vector $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots \mathrm{d}$ of base calls and their associated quality scores, where $\mathrm{d}$ is total read depth.
We want to use $\mathbf{x}$ to select between competing hypotheses;
$$
  \begin{array}{l}
    \mathbf{H_0}:\quad \textrm{Alt allele} = m;\quad\nu = 0\\
    \mathbf{H_1}:\quad \textrm{Alt allele} = m;\quad\nu = \hat{f},
  \end{array}
$$
where $\nu$ is the variant allele frequency, $\hat{f}$ is the maximum likelihood estimate of $\nu$ given data $\mathbf{x}$, i.e. the ratio of the count of variant reads and total read depth, and $m$ is any of the 3 possible alternate non-reference bases.
The posterior probability of a given hypothesis, $P(m,\nu)$, is the product of the likelihood of the data given the hypothesis and the prior probability of the hypothesis. 
Assuming that reads are independent, we have

\begin{equation}
  \label{eqn:1}
$$
  P(m,\nu) = p(m,\nu) \cdot \prod_{i=1}^{\mathrm{d}} \textrm{f}_{m,\nu}(x_i),
$$
\end{equation}
where the density $\textrm{f}_{m,\nu}(x_i)$ can take several forms.

@@

\subsection{Estimation of the mutation profile}
There are many aspects of genomic architecture which can affect the somatic mutation rate at multiple scales \cite{Buisson2019}.
Here we focus on a small-scale feature, the tri-nucleotide context, which has been shown to have a strong effect on the prior probability of mutation \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
The tri-nucleotide context at a particular genomic site is made up of the identity of the reference base and the identity of the 3` and 5` flanking bases.
There are four possible bases {\{A,C,T,G\}} 3` and 5` of the site, leading to 16 possible flanking base combinations.
Folding the central bases to the pyrimidines leaves 2 possible reference bases at the site {\{C,T\}}, resulting in $2 \cdot 16 = 32$ possible tri-nucleotide contexts.
We define a mutation context $C$ indexed by $i=\{1 \dots 32\}$, to be the tri-nucleotide context of a particular site in the genome. 
For each context, a mutation can be to any of the three possible alternate alleles {\{A,(C/T),G\}}. 
We can thus define $\mathrm{S}$, the substitution type, indexed by $i=\{1 \dots 32\}$ contexts and $j = \{1 \dots 3\}$ alternate bases, resulting in 96 possible substitution types.
Using the substitution type $\mathrm{S}$, we can rewrite equation \ref{eqn:full_prior} as

\begin{equation}
  \label{eqn:detailed_prior}
  $$
  \begin{array}{rcc}
  \mathrm{p}(\mathrm{S}) &=&  \mathrm{p}(m =j, M | C = i) \\
                            &=& \mathrm{p}(m = j \mid C = i) \mathrm{p}(C = i \mid M)\frac{p(M)}{p(C = i)},
  \end{array}
  $$
\end{equation}
where $\mathrm{p}(\mathrm{S})$ is the probability that a substitution of type $S$ has occurred.

\begin{figure}[t]
  \begin{center}
  \includegraphics{figures/signature_figure.pdf}
  \end{center}
  \caption{Mutation profiles. 
  \textbf{(a)} The observed mutation profile of a breast cancer sample from \cite{Shi2018}.
  \textbf{(b)} The observed mutation profile of the AML 31 primary tumor from \cite{Griffith2015}.
  \textbf{(c)} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
  \textbf{(d)} Equal proportions of signatures 1, 4, \& 5.
  \textbf{(e)} Equal proportions of signatures 1, 3, \& 5.
  KL divergence between the estimated profile ($\boldsymbol{\pi}$) and the simulated mutation profile for \textbf{(f)} Whole exomes and \textbf{(g)} whole genomes.
  The estimated mutation profile converges quickly to the target as the number of mutations evaluated rises.
  }
  \label{NAR-sigfig}
  \end{figure}

Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.
The mutation profiles in Figure \ref{NAR-sigfig} can be described compactly by the vector $\boldsymbol{\pi}$ where each element of $\boldsymbol{\pi}$ represents the proportion of all observed tumor mutations of substitution type $\mathrm{S}_{i,j}$.
The distribution of observed substitutions $ \mathrm{p}(\mathrm{S}) $, is multinomial with parameter $ \boldsymbol{ \pi } = \{\pi_{i,j}\} $.
In a tumor with a very large number of verified mutations $\boldsymbol{\pi}$ could be estimated directly from the observed mutation profile, but in practice we find that many elements of $\boldsymbol{\pi}$ will have $0$ or a very small number of mutations present in the tumor.
As a result we need to model $p(S)$ as dirichlet - multinomial with pseudo-count hyper-parameter $\boldsymbol{\alpha}$,

$$
\begin{aligned}
  \boldsymbol{\pi} \mid \boldsymbol{\alpha} &\sim \textrm{Dirichlet}(\boldsymbol{\alpha}) \\
  S \mid \boldsymbol{\pi} & \sim \textrm{Multinomial}(\boldsymbol{\pi}).
\end{aligned}
$$
In \batcave we use the symmetric non-informative hyper-parameter $\boldsymbol{\alpha} = \boldsymbol{1}$, so a-priori a mutation is equally likely in any context.

@@

Let $\mathrm{D}$ be the set of high confidence variant calls, and $c \in \mathrm{D}$ be the context of each mutation in $\mathrm{D}$.
The posterior distribution of $\boldsymbol{\pi}$ is then $\mathrm{p}(\boldsymbol{\pi} \mid \mathrm{D}) \sim \textrm{Dirichlet}(\boldsymbol{\alpha^{\prime}})$ where
  $$
    \alpha^{\prime}_{i,j} = \alpha_{i,j} + \sum\limits_{c \in \mathrm{D}} \mathrm{I}\{c = c_{i,}\}.
  $$
Returning to equation \ref{eqn:detailed_prior}, we can compute the posterior predictive probability $p(C_{i} \mid M,\mathrm{D})$ as
\begin{equation}
  \label{eqn:post_pred}
  $$
  p(C = C_{i} \mid M,\mathrm{D}) = \frac{\sum_{j}\alpha^{\prime}_{i,j}}{\sum_{i,j}\alpha^{\prime}_{i,j}}.
  $$
\end{equation}
The posterior probability of mutation to allele $m$ given that a mutation has occurred in context $C_{i}$, $p(m = j \mid C_{i},\mathrm{D})$ is then
\begin{equation}
  \label{eqn:to_allele}
  $$
   p(m = j \mid C_{i},\mathrm{D}) = \frac{\alpha^{\prime}_{i,j}}{\sum_{j} \alpha^{\prime}_{i,j}}.
  $$
\end{equation}
The remaining context dependent component in equation \ref{eqn:full_prior} is $p(C = C_{i})$, the prior probability of seeing a particular tri-nucleotide context.
We compute this as simply the proportion of all tri-nucleotide contexts with the context $C_{i}$ present in the genome.
\bkmcomment{I have a pre-computed table in batcaver for wes and wgs. Should I add that bit here?}
